Using Nature Inspired Algorithms for Feature Selection in Transaction Fraud Detection

### 2020-05-31 paper/main.tex
Add finishing touches

\begin{document}
%
\title{Transactions Fraud Detection using Machine Learning and Nature Inspired Algorithms}
%
\titlerunning{Transactions Fraud Detection using ML and NIA}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Peter Mačinec \and Timotej Zaťko}
%
\authorrunning{Peter Mačinec, Timotej Zaťko}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Faculty of Informatics and Information Technologies,\\Slovak University of Technology, Bratislava}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Credit card payments are still more and more preferred because of convenience. People are using credit cards to pay not only at stores, but also in e-shops. However, with the rise of credit cards usage, suitable environment for frauds in transactions arises. The data of transactions are usually collected by the banks, that indicates usage of data-mining approaches to prevent frauds. The performance of model is based on several sub-steps and decisions, like data analysis, preprocessing, feature selection and hyperparameter optimization. In the area of transactions fraud detection, datasets usually contain a lot of features (several hundreds) that are however mostly anonymized. Thus we decided to focus on improving model mainly by selecting appropriate features using nature inspired algorithms. Nature inspired algorithms proved to be powerful methods in such optimization tasks. Experiments showed that proper combination of machine learning model with feature selection using nature inspired algorithms has potential to help to prevent transaction frauds.

\keywords{transactions fraud detection \and machine learning classification \and nature inspired algorithms \and feature selection \and bat algorithm.}
\end{abstract}
%
%
%
\section{Introduction}

Payments with credit cards are nowadays still more preferred over using cash. Using credit card is not only more comfortable for people, but even also more safe than carrying cash in wallet when it comes to higher amount of money. However, number of transaction frauds is arising alongside with usage of credit cards. Number of transactions and ability to obtain data from them indicate the need of automatic detection of suspicious payments.

Data of performed transactions via credit cards are naturally collected by companies and banks to produce statistics or investigate frauds. Therefore, transactions fraud detection can be interpreted as data mining problem, concretely binary classification.

In this paper, we propose novel method for detection frauds in transactions using aspects of data analysis, machine learning and nature inspired algorithms. The basis of our method lies in training machine learning model on best features selected by nature inspired algorithm. More nature inspired algorithms are compared to choose the best one for this problem. Finally, nature inspired algorithms in combination with machine learning proved to be very efficient way for detecting frauds in transactions, overperforming common methods in this area.
% TODO ten overperforming treba nejak doladit mozno

% Required: Úvod / Introduction - aká úloha sa ide riešiť, prečo je dobré ju riešiť

% TODO: co robime, tj, nacrtnut problem atd
% TODO: spomenut, datasety s velkym mnozstvom features, je potrebna feature selection
% TODO: spomenut, ze robi sa to takto takto, ale su aj nejako prirodou inspirovane algoritmy, ktore maju taketo vyhody
% TODO: preco? napr. mzoe chybat domenovat znalost (features) kvoli anonymizacii - nas pripad

@@

@@

Majority of datasets available for data science research of transactions fraud detection have the same characteristics - highly imbalanced data, a lot of features and majority of features are anonymized.

Our method will be trained and evaluated on dataset from Kaggle competition EEE-CIS Fraud Detection\footnote{https://www.kaggle.com/c/ieee-fraud-detection/data}. As usual in problems of transactions fraud detection, a lot of features of different types are available - 434 features describing demography, credit card, transaction itself, etc. Majority of features are anonymized or the meaning of them is not clear. In this case, we must be careful to avoid labels leak from some features and also in interpreting data analysis or models result when talking about anonymized features. This dataset contains almost 600k samples, that is appropriate for machine learning algorithms. However, data are highly imbalanced - classes distribution can bee seen at figure~\ref{fig:classes}. When training and evaluating models, one should be careful when data are imbalanced.

@@

We propose to use nature-inspired algorithms for feature selection. We will try several nature-inspired algorithms and compare them to find out which one is most suitable for our problem. We decided to use the following algorithms: Firefly Algorithm (FA) \cite{fister2013comprehensive}, Cuckoo Search (CS) \cite{yang2009cuckoo}, Bat Algorithm (BA) \cite{yang2010new}, Flower Pollination Algorithm (FPA) \cite{yang2012flower}, Grey Wolf Optimizer (GWO) \cite{Mirjalili_Mirjalili_Lewis_2014} since they were already used for feature selection in related works or they performed extraordinarily in other types of tasks.

@@

\begin{figure}[ht]
	\begin{center}
	    \includegraphics[width=6cm]{figures/projection.png}
    \end{center}
	\caption{Projection from the vector of floats which is being optimized by nature-inspired algorithms to the binary vector representing wether feature is selected or not.}
	\label{fig:projection}
\end{figure}

@@

We did several experiemnts in which we compared the following algorithms with these parameters:

\begin{itemize}
	\item Firefly Algorithm (alpha = 1, betamin = 1, gamma = 2)
	\item Cuckoo Search (pa = 0.2, alpha = 0.5)
	\item Bat Algorithm (A = 0.5, r = 0.5, Qmin = 0.0, Qmax = 2.0)
	\item Flower Pollination Algorithm (p = 0.35, beta = 1.5)
	\item Grey Wolf Optimizer
\end{itemize}

As a baseline models we have chosen the model with all features and model with selected features from \textit{recursive feature selection with cross-validation}.

@@

We chose a \textit{decision tree classifier} as a model on which we will try to find the optimal features. We kept the default hyper-parameters for this classifier set by scikit-learn library.

@@

For each algorithm, we tried several population sizes (10, 20, and 50), for which we did 5 runs of each algorithm. We trained exactly 50 generations. For both, recursive feature selection and feature selection using nature-inspired algorithms we did 5 fold cross-validation. We did not experiment with algorithm parameters and we used their defaults, as we mentioned in the previous section.

@@

When comparing the population sizes, all of the algorithms achieved the best results when the population size was highest (50). The only exception is Cuckoo Search, which found the most optimal solution with population size of 10, however the difference negligible. Table \ref{tab:nia_comparison}. shows the algorithm scores for each population. Figure \ref{fig:nia_train_score_by_algorithm}. shows the differences in scores between population size of 10 and 50.

\begin{figure}
    \centering
    \subfloat{{ 
        \includegraphics[width=5.5cm]{figures/nia_train_score_by_algorithm_10}
    }}
    \qquad
    \subfloat{{
        \includegraphics[width=5.5cm]{figures/nia_train_score_by_algorithm_50}
    }}
    \caption{Scores of the algorithms by the population size. The scores are slightly higher for population size of 50.}
    \label{fig:nia_train_score_by_algorithm}
\end{figure}

\begin{table}[]
	\begin{center}
		\begin{tabular}{|l|rrrrr|}
			\hline
			\textbf{Population size \textbackslash Score} & \multicolumn{1}{l}{\textbf{FA}} & \multicolumn{1}{l}{\textbf{CS}} & \multicolumn{1}{l}{\textbf{BA}} & \multicolumn{1}{l}{\textbf{FPA}} & \multicolumn{1}{l|}{\textbf{GWO}} \\ \hline
			\textbf{10}                                             & 0.753                           & \textbf{0.752}                           & 0.759                           & 0.754                            & 0.776                             \\
			\textbf{20}                                             & 0.757                           & 0.746                           & 0.755                           & 0.750                            & 0.771                             \\
			\textbf{50}                                             & \textbf{0.761}                           & 0.751                           & \textbf{0.761}                           & \textbf{0.756}                            & \textbf{0.782}                             \\ \hline
		\end{tabular}
	\end{center}
	\begin{center}
		\caption{Highest train scores (ROC AUC) of nature inspired algorithms by population size. We may notice that Grey Wolf Algorithm (GWO) outperformed all of the other algorithms.} \label{tab:nia_comparison}
	\end{center}
\end{table}

All of the nature-inspired algorithms perform roughly the same, except the Grey Wolf Optimizer which outperformed the others (Figure \ref{fig:nia_train_score_by_algorithm}). These differences in results between GWO and other algorithms are significant (there is a sigificant difference in means, t-test; $p-value=0.012$). Looking for the explanation, we found out that only Grey Wolf Algorithm and Bat Algorithm converged during the optimization run. Figure \ref{fig:nia_scores_by_run} shows the optimization progress of the Firefly Algorithm and Grey Wolf Optimizer, Firefly Algorithm did not converge. If we fine-tuned the parameters of other algorithms, they might converge as well. The fact that GWO does not have any parameters (except the population size) might be an advantage for it in our experiments.

@@

\begin{figure}
    \centering
    \subfloat[Firefly Algorithm]{{ 
        \includegraphics[width=5.5cm]{figures/fa_scores_by_run}
    }}
    \qquad
    \subfloat[Grey Wolf Optimizer]{{
        \includegraphics[width=5.5cm]{figures/gwo_scores_by_run}
    }}
    \caption{Optimization progress for 5 runs of Grey Wolf Optimizer and Firefly Algorithm. The Firefly Algorithm did not really converge.}
    \label{fig:nia_scores_by_run}
\end{figure}

@@

\begin{table}[]
	\begin{center}
		\begin{tabular}{|lrrr|}
			\hline
			\textbf{Feature selection type} & \textbf{Train score} & \textbf{Test score} & \textbf{Selected features} \\ \hline
			All features                    & N/A                  & 0.713               & 240                                  \\
			Recursive feature elimination   & 0.747                & 0.719               & 112                                  \\
			Firefly Algorithm               & 0.761                & 0.663               & 114                                  \\
			Cuckoo Search                   & 0.753                & 0.718               & 106                                  \\
			Bat Algorithm                   & 0.761                & 0.689               & 125                                  \\
			Flower Pollination Algorithm    & 0.756                & 0.706               & 113                                  \\
			\textbf{Grey Wolf Optimizer}             & \textbf{0.782}                & \textbf{0.733}               & \textbf{53}                                   \\ \hline
		\end{tabular}
	\end{center}
	\begin{center}
		\caption{Comparison of the baseline models with nature-inspired algorithms. The train score stands for the best score from the optimization. All of the nature-inspired algorithms (except Grey Wolf Optimizer) achieved a better train score than recursive feature elimination, however, they overfit the model which resulted in a worse test scores.} \label{tab:nia_baseline_comparison}
	\end{center}
\end{table}

\begin{figure}[ht]
	\begin{center}
	    \includegraphics[width=5.5cm]{figures/nia_selected_features_by_algorithm.png}
    \end{center}
	\caption{Number of selected features by algorithm - Grey Wolf Algorithm selected by far the least number of features.}
	\label{fig:nia_selected_features_by_algorithm}
\end{figure}

>>>>

\begin{document}
%
\title{Using Nature Inspired Algorithms for Feature Selection in Transaction Fraud Detection}
%
\titlerunning{Using NIA for Feature Selection in Transaction Fraud Detection}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Peter Mačinec \and Timotej Zaťko}
%
\authorrunning{Peter Mačinec, Timotej Zaťko}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Faculty of Informatics and Information Technologies,\\Slovak University of Technology, Bratislava}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Credit card payments are still more and more preferred because of convenience. People are using credit cards to pay not only at stores but also in e-shops. However, with the rise of credit card usage, a suitable environment for fraud in transactions arises. The data of transactions are usually collected by the banks, which indicates the usage of data-mining approaches to prevent frauds. The performance of the model is based on several sub-steps and decisions, like data analysis, preprocessing, feature selection, and hyperparameter optimization. In the area of transaction fraud detection, datasets usually contain a lot of features (several hundred) that are however mostly anonymized. Thus we decided to focus on improving model mainly by selecting appropriate features using nature-inspired algorithms. Nature-inspired algorithms proved to be powerful methods in such optimization tasks. Experiments showed that a proper combination of machine learning model with feature selection using nature-inspired algorithms has the potential to help to prevent transaction frauds. Out of 5 nature-inspired algorithms, the Grey Wolf Optimizer achieved the best results. It improved the model by 1.4\% (ROC AUC) while selecting half of the features in comparison with a recursive feature elimination.

\keywords{transaction fraud detection \and machine learning classification \and nature inspired algorithms \and feature selection \and bat algorithm.}
\end{abstract}
%
%
%
\section{Introduction}

Payments with credit cards are nowadays still more preferred over using cash. Using a credit card is not only more comfortable for people, but even also more safe than carrying cash in wallet when it comes to a higher amount of money. However, the number of transaction frauds is arising alongside with usage of credit cards. The number of transactions and the ability to obtain data from them indicates the need for automatic detection of suspicious payments.

Data of performed transactions via credit cards are naturally collected by companies and banks to produce statistics or investigate frauds. Therefore, transaction fraud detection can be interpreted as a data mining problem, concretely binary classification. The majority of existing datasets of transaction frauds contain a lot of features, however, most of them are usually anonymized. Thus, the model will rely also on proper feature selection. In such cases, where feature selection is a very important part of the problem, powerful metaheuristic search methods like nature-inspired algorithms can be used.

In this paper, we propose a novel method for detection frauds in transactions using aspects of data analysis, machine learning, and nature-inspired algorithms. The basis of our method lies in the training machine learning model on the best features selected by the nature-inspired algorithm. More nature-inspired algorithms are compared to choose the best one for this problem.

@@

The number of transaction frauds is arising alongside with usage of credit cards. To help prevent transaction frauds, common data-mining techniques are available. However, data are in most cases very specific and there is no method suitable to all problems. Addressing that, a comparative study of various machine learning techniques, mostly neural networks, for credit card fraud detection was performed~\cite{Sadgali2019}. Common algorithms like Support Vector Machines~\cite{Wiese2009} or Random Forest~\cite{Xuan2018} were also examined. Another and very specific way to detect frauds is by using Genetic programming~\cite{Assis2013}. Genetic programming has proven to be very effective for this problem, achieving significant improvements (up to almost 17\% compared to corporation baseline).

Each transaction can be viewed separately as one unit, but also whole transaction history can be considered. To create a model that is resistant to changes in transactions, modeling time series using Long-Short Term Memory (LSTM) recurrent neural networks seems to be the right way~\cite{Wiese2009}.

Publicly available datasets for transaction fraud detection usually contain a lot of features. It is necessary to put special effort into the feature selection phase in such cases. Nature-inspired algorithms proved to be very powerful for feature selection problems in previous research. When using metaheuristic approaches like nature-inspired algorithms, feature selection problem can be interpreted as combinatorial optimization problem~\cite{Wang2016}.

Rough sets are popular when using nature-inspired algorithms for feature selection (more about rough sets theory used in data mining in~\cite{Slimani2013}). Rough sets were used in combination with Firefly Algorithm~\cite{Banati2011}, Particle Swarm Optimization~\cite{Wang2007}, or Bat Algorithm~\cite{Emary2014}. The majority of works used multiple datasets for their experiments to prove nature-inspired algorithms power.

There are also other surveys focused mainly on the usage of nature-inspired algorithms for feature selection~\cite{Kauser2018,Wang2016,Yang2020}. Further algorithms were researched, like Artificial Bees Colony, Cuckoo Search, Ant Colony Optimization, or Genetic Algorithms.

@@

The majority of datasets available for data science research of transaction fraud detection have the same characteristics - highly imbalanced data, a lot of features, and the majority of features are anonymized.

Our method will be trained and evaluated on a dataset from Kaggle competition EEE-CIS Fraud Detection\footnote{https://www.kaggle.com/c/ieee-fraud-detection/data}. As usual in problems of transaction fraud detection, a lot of features of different types are available - 434 features describing demography, credit card, the transaction itself, etc. The majority of features are anonymized or the meaning of them is not clear. In this case, we must be careful to avoid labels leak from some features and also in interpreting data analysis or models result when talking about anonymized features. This dataset contains almost 600k samples, that is appropriate for machine learning algorithms. However, data are highly imbalanced - class distribution can be seen in figure~\ref{fig:classes}. When training and evaluating models, one should be careful when data are imbalanced.

We cannot handpick the features, because the data are anonymized so we lack the domain knowledge. Hence we need to apply some methods for feature selection. Also, manually hand-picking for that large amount of features would be nearly impossible. Another motivation for feature selection is that it reduces model complexity and improves its interpretability.

@@

% 440 features at the end

To improve a machine learning models for the transaction fraud detection in a way of interpretability and complexity, we propose to use nature-inspired algorithms for feature selection. We will try several nature-inspired algorithms and compare them to find out which one is most suitable for our problem. We decided to use the following algorithms: Firefly Algorithm (FA) \cite{fister2013comprehensive}, Cuckoo Search (CS) \cite{yang2009cuckoo}, Bat Algorithm (BA) \cite{yang2010new}, Flower Pollination Algorithm (FPA) \cite{yang2012flower}, Grey Wolf Optimizer (GWO) \cite{Mirjalili_Mirjalili_Lewis_2014} since they were already used for feature selection in related works or they performed extraordinarily in other types of tasks.

@@

\begin{figure}[ht]
	\begin{center}
	    \includegraphics[width=8cm]{figures/projection.png}
    \end{center}
	\caption{Projection from the vector of floats which is being optimized by nature-inspired algorithms to the binary vector representing wether feature is selected or not.}
	\label{fig:projection}
\end{figure}

@@

We did several experiemnts in which we compared the following algorithms and their parameters:

\begin{itemize}
	\item Firefly Algorithm ($\alpha = 1$, $\beta_{min} = 1$, $\gamma = 2$)
	\item Cuckoo Search ($p_a = 0.2$, $\alpha = 0.5$)
	\item Bat Algorithm ($A = 0.5$, $r = 0.5$, $Q_{min} = 0.0$, $Q_{max} = 2.0$)
	\item Flower Pollination Algorithm ($p = 0.35$, $\beta = 1.5$)
	\item Grey Wolf Optimizer
\end{itemize}

We will use the algorithm default parameters as they are implemented in \textit{NiaPy} \cite{NiaPyJOSS2018} library based on the original papers. As a baseline models we have chosen the model with all features and model with selected features from \textit{recursive feature elimination with cross-validation}.

@@

We chose a \textit{decision tree classifier} as a model on which we will try to find the optimal features. Based on our experiments it achieved the best results considering also the training speed. We kept the default hyper-parameters for this classifier set by scikit-learn library.

@@

For each algorithm, we tried several population sizes (10, 20, and 50), for which we did 5 runs of each algorithm. We trained exactly 50 generations. For both, recursive feature elimination and feature selection using nature-inspired algorithms we did 5 fold cross-validation.

@@

When comparing the population sizes, all of the algorithms achieved the best results when the population size was highest (50). The only exception is Cuckoo Search, which found the most optimal solution with population size of 10, however the difference negligible. Table \ref{tab:nia_comparison}. shows the algorithm scores for each population. Figure \ref{fig:nia_train_score_by_algorithm_1}. \& \ref{fig:nia_train_score_by_algorithm_2}.shows the differences in scores between population size of 10 and 50.

% \begin{figure}
%     \centering
%     \subfloat{{ 
%         \includegraphics[width=5.5cm]{figures/nia_train_score_by_algorithm_10}
%     }}
%     \qquad
%     \subfloat{{
%         \includegraphics[width=5.5cm]{figures/nia_train_score_by_algorithm_50}
%     }}
%     \caption{Scores of the algorithms by the population size. The scores are slightly higher for population size of 50.}
%     \label{fig:nia_train_score_by_algorithm}
% \end{figure}

\begin{figure}
    \centering
    \includegraphics[width=11cm]{figures/nia_train_score_by_algorithm_10}
    \caption{Scores of the nature-inspired algorithms for population size of 10.}
    \label{fig:nia_train_score_by_algorithm_1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=11cm]{figures/nia_train_score_by_algorithm_10}
    \caption{Scores of the nature-inspired algorithms for population size of 50.}
    \label{fig:nia_train_score_by_algorithm_2}
\end{figure}

\def\arraystretch{1.5}
\begin{table}
	\begin{center}
		\caption{Highest train scores (ROC AUC) of nature inspired algorithms by population size. We may notice that Grey Wolf Algorithm (GWO) outperformed all of the other algorithms.} \label{tab:nia_comparison}
	\end{center}
	\begin{center}
		\begin{tabular}{|l|r|r|r|r|r|}
			\hline
			\textbf{Population size \textbackslash Score} & \multicolumn{1}{r|}{\textbf{FA}} & \multicolumn{1}{r|}{\textbf{CS}} & \multicolumn{1}{r|}{\textbf{BA}} & \multicolumn{1}{r|}{\textbf{FPA}} & \multicolumn{1}{r|}{\textbf{GWO}} \\ \hline
			\textbf{10}                                             & 0.753                           & \textbf{0.752}                           & 0.759                           & 0.754                            & 0.776                             \\
			\textbf{20}                                             & 0.757                           & 0.746                           & 0.755                           & 0.750                            & 0.771                             \\
			\textbf{50}                                             & \textbf{0.761}                           & 0.751                           & \textbf{0.761}                           & \textbf{0.756}                            & \textbf{0.782}                             \\ \hline
		\end{tabular}
	\end{center}
\end{table}

All of the nature-inspired algorithms perform roughly the same, except the Grey Wolf Optimizer which outperformed the others (Figure \ref{fig:nia_train_score_by_algorithm}). These differences in results between GWO and other algorithms are significant (there is a sigificant difference in means, t-test; $p-value=0.012$). Looking for the explanation, we found out that only Grey Wolf Algorithm and Bat Algorithm converged during the optimization run. Figure \ref{fig:nia_scores_by_run_1}. \& \ref{fig:nia_scores_by_run_2}. shows the optimization progress of the Firefly Algorithm and Grey Wolf Optimizer, Firefly Algorithm did not converge. If we fine-tuned the parameters of other algorithms, they might converge as well. The fact that GWO does not have any parameters (except the population size) might be an advantage for it in our experiments.

@@

\begin{figure}[htp]
	\centering
	\includegraphics[clip, width=10.5cm]{figures/gwo_scores_by_run}
    \caption{Optimization progress for 5 runs of Grey Wolf Optimizer.}
    \label{fig:nia_scores_by_run_1}
\end{figure}

\begin{figure}[htp]
	\centering
	\includegraphics[clip, width=10.5cm]{figures/fa_scores_by_run}
    \caption{Optimization progress for 5 runs of Firefly Algorithm. The optimization process did not converge.}
    \label{fig:nia_scores_by_run_2}
\end{figure}

@@

\begin{table}[]
	\begin{center}
		\caption{Comparison of the baseline models with nature-inspired algorithms. The train score stands for the best score from the optimization. All of the nature-inspired algorithms (except Grey Wolf Optimizer) achieved a better train score than recursive feature elimination, however, they overfit the model which resulted in a worse test scores.} \label{tab:nia_baseline_comparison}
	\end{center}
	\begin{center}
		\begin{tabular}{|l|r|r|r|}
			\hline
			\textbf{Feature selection type} & \textbf{Train score} & \textbf{Test score} & \textbf{Selected features} \\ \hline
			All features                    & N/A                  & 0.713               & 240                                  \\
			Recursive feature elimination   & 0.747                & 0.719               & 112                                  \\
			Firefly Algorithm               & 0.761                & 0.663               & 114                                  \\
			Cuckoo Search                   & 0.753                & 0.718               & 106                                  \\
			Bat Algorithm                   & 0.761                & 0.689               & 125                                  \\
			Flower Pollination Algorithm    & 0.756                & 0.706               & 113                                  \\
			\textbf{Grey Wolf Optimizer}             & \textbf{0.782}                & \textbf{0.733}               & \textbf{53}                                   \\ \hline
		\end{tabular}
	\end{center}
\end{table}

\begin{figure}[ht]
	\begin{center}
	    \includegraphics[width=10cm]{figures/nia_selected_features_by_algorithm.png}
    \end{center}
	\caption{Number of selected features by algorithm - Grey Wolf Algorithm selected by far the least number of features.}
	\label{fig:nia_selected_features_by_algorithm}
\end{figure}
